Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-08-10 text/sections/01-introduction.tex
adjusted writing

There are various margins through which funding cuts can impact a university's faculty.
A university's level of resources affect the ability that departments can hire new professors, retain their current faculty, or provide retirement incentives.
So that it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
Yet, it is a priori unclear whether/how much each of these margins adjusts to funding cuts, as these decision depend on how the university administrations view the productivity of different inputs, plus the degree of wage and employment stickiness over time.

@@

I use data from the Integrated Postsecondary Education Data System (IPEDS) to study the impact of changes in state funding for US public univeristies 1990--2017.
I combine these data with panel data for every public university faculty member in the state of Illinois for 2010--2021 (Illinois Board of Hdigher Education Database, IBHED), to measure how faculty are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), then do they adjust the number of and composition of faculty they employ, and how does this affect individual faculty already at the university?

@@

Falls in state funding affect faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a (funding per student) cut of 10\% leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the year of the funding shock. 
Over the same time period, state funding per student fell by around 35\%, count of (tenure-tack/tenured) professors per student fell by 9\%, and lecturers per student increased by 99\%, so that these results show that falls in state funding explain around 53\% of the fall in professors per student
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

>>>>

There are various margins through which funding cuts can impact a university's faculty.
A university's level of resources affect the ability that departments can hire new professors, retain their current faculty, or provide retirement incentives, so it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
Yet, it is a priori unclear whether/how much each of these margins adjusts to funding cuts, as these decision depend on how the university administrations view the productivity of different inputs, plus the degree of wage and employment stickiness over time.

@@

I use data from the \cite{ipeds} to study the impact of changes in state funding for US public universities 1990--2017.
I combine these data with panel data for every public university faculty member in the state of Illinois for 2010--2021 \citep{ibhed}, to measure how faculty are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), do they change the composition of faculty they employ?
If so, which positions do universities substitute away from, or towards, and do these funding cuts affect the faculty members themselves?

@@

Falls in state funding affect faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a funding cut of 10\% per student leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the year of the funding shock. 
Over the same time period, state funding per student fell by around 35\%, count of (tenure-tack/tenured) professors per student fell by 9\%, and lecturers per student increased by 99\%, so that these results show that falls in state funding explain around 53\% of the fall in professors per student
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

### 2023-08-10 text/sections/03-concepts.tex
adjusted writing

Employment composition changes likely arise via differences in hiring rate across levels of professor seniority.
Universities hire new professors most years, either to expand their departments or to replace leaving professors, but not hire at the same rate for each level of seniority.
The number of professors hired in each year is usually highest among non-tenured adjunct faculty, as these instructors are mostly on short-term or contingent contracts, so  are less costly to hire (or to let go) in response to yearly changes.
Tenure-track assistant professors are mostly hired with a four to six year contract, and formal agreement for tenure consideration at the end of this term.
Lastly, tenured professors have successfully secured a full-time appointment at their university with no expiration date; this position has the lowest yearly hiring rate among most universities.
Similarly, there is highest turn-over (including leaving the university) among lecturers, and lowest among full professors.
The differences in yearly hiring and turnover rate mean that if hiring is affected by state funding, then there will be faculty composition change in the years following budget cuts.

@@

Similarly, its is not clear when the effects of funding short-falls will be realised.
University departments often employ faculty on multiple-year contracts (e.g., the tenure contract), and individual professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for university funding to change the faculty composition, or trickle down to faculty.
For example, university departments coordinate hiring in year-long cycles, so that a budget short-fall in year $t$ may have no effect on the in-progress hiring committee and resulting hiring decisions, yet the hiring cycle in year $t+1$ may be postponed or cancelled in response.
As such, this paper investigates the dynamics of state funding shocks' effects on faculty, for the years following a university funding shock.

>>>>

Employment composition changes likely arise via differences in hiring rate across levels of professor seniority.
Universities hire new professors most years, either to expand their departments or to replace leaving professors, but do not hire at the same rate for each level of seniority.
The yearly hiring is usually highest among non-tenured adjunct faculty, where universities hire instructors often on short-term or contingent contracts, so that lecturers are less costly to hire (or to let go) in response to yearly changes.
Tenure-track assistant professors are mostly hired with a four to six year contract, and formal agreement for tenure consideration at the end of this term.
Lastly, tenured professors have successfully secured a full-time appointment at their university with no expiration date; this position has the lowest yearly hiring rate among most universities.
Similarly, there is highest turn-over (including leaving the university) among lecturers, and lowest among full professors.
The differences in yearly hiring and turnover rate mean that if hiring is affected by state funding, then there will be faculty composition change in the years following budget cuts.

@@

Similarly, its is not clear when the effects of funding short-falls will be realised.
University departments often employ faculty on multiple-year contracts (e.g., the tenure contract), and individual professors' decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for university funding to change the faculty composition, or trickle down to faculty.
For example, university departments coordinate hiring in annual cycles, so that a budget short-fall in year $t$ may have no effect on the in-progress hiring committee and resulting hiring decisions, yet the hiring cycle in year $t+1$ may be postponed or cancelled in response.
As such, this paper investigates the dynamics of state funding shocks' effects on faculty, for the years following a university funding shock.

### 2023-08-10 text/sections/04-empirical.tex
adjusted writing

The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects;
Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
Columns (1) shows that a funding shock of \$10 (per student in the entire state) is associated with \$11.76 less state funding per student at the univeristy, with the corresponding -10\% funding shock leads to -9.77\%
less funding (coumn 1, panel B), with similar estimates with and without including fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shock, likely thanks to systematic differences in universities unaccounted for without fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider (or bad control).
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that the uncertainty in including tuition revenue as a possible bad control for the level of state funding does not matter thanks to the inclusion of fixed effects.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

>>>>

The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.
\autoref{tab:firststage-reg} Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
Column (1) shows that a funding shock of \$10 (per student in the entire state) is associated with \$11.76 less state funding per student at the university, with the corresponding -10\% funding shock leads to -9.77\%
less funding (column 1, panel B), with similar estimates with and without including fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shock, likely thanks to systematic differences in universities unaccounted for without fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider (or bad control).
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that the uncertainty in including tuition revenue as a possible bad control for the level of state funding does not matter thanks to the inclusion of fixed effects.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

### 2023-08-10 text/sections/10-appendix.tex
adjusted writing

\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Correlation Between State Funding Shock and Public University State Funding in Surrounding Years.}
    \includegraphics[width=\textwidth]{figures/lag-firststage.png}
    \label{fig:lag-firststage}
    \justify
    \footnotesize
    \textbf{Note}:
    This figure shows the correlation between state funding in year $t+k$ with the funding shock in year $t$ for a university, where $k = -7, \hdots, 5$ are the years on the x-axis.
    This shows that state funding and the funding shock have dynamic correaltion, so that dynamic effects must be estimated by localc projects -- and not simple OLS or 2SLS.
    The estimates are of \eqref{eqn:firststage}, calculated with IPEDS data, separatly for each year relative to initial shock, using the $\log-\log$ specification, including fixed effects for university $+$ year, and clustering standard errors by university $+$ year.
\end{figure}

@@

Yearly variation is not observed here, so that only the aggregate level, for 180 universities, can be considered.
\begin{table}[H]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \label{tab:hiring-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}: 
    This table show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on the number of faculty hires (per student) total for 2011--2021 at US public universities, using the funding shock to instrument for state funding.
    Each observation is a university--year at a public university, where funding data come from IPEDS and faculty count total for 2011--2021 from \citep{wapman2022quantifying}.
    The panels shows the effect of a $1$\% change in state funding per student at the university (total for 2011--2021) on the number of new faculty hires by gender (and all).
    Standard errors are clustered at the university-year level, and university $+$ year fixed effects are included through--out.
\end{table}

@@

The implied marginal rate of substitution between lecturers and assistant professors is estimated as -3.26 (standard error 0.50), based on based on 10,000 bootstrap samples.
This means that public universities increased their number lecturers per student by 3.26\% when they decreased their count of assistant professors, on average and subject to the changes in state funding they experienced 1990--2017.
Between lecturers and full professors the rate of subsitution is -3.19 (0.34), which implies that universities subsitute between lecturers and full professors in the same way.
Between assistant and full professors the rate of substitution is 0.99 (0.11), which intuitively implies that universities treated assistant and full professors (i.e., those before and after tenure in the tenure system) as complements.

>>>>

\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Correlation Between State Funding Shock and Public University State Funding in Surrounding Years.}
    \includegraphics[width=\textwidth]{figures/lag-firststage.png}
    \label{fig:lag-firststage}
    \justify
    \footnotesize
    \textbf{Note}:
    This figure shows the correlation between state funding in year $t+k$ with the funding shock in year $t$ for a university, where $k = -7, \hdots, 5$ are the years on the x-axis.
    This shows that state funding and the funding shock have dynamic correlation, so that dynamic effects must be estimated by local projections -- and not simple OLS or 2SLS.
    The estimates are of \eqref{eqn:firststage}, calculated with IPEDS data, separately for each year relative to initial shock, using the $\log-\log$ specification, including fixed effects for university $+$ year, and clustering standard errors by university $+$ year.
\end{figure}

@@

Yearly variation is not observed here, so that only the aggregate level, for 180 universities, can be considered.
\begin{table}[H]
    \singlespacing
    \centering
    \caption{OLS and 2SLS Estimates for University Faculty Hiring, Total for 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/hiring-shock-reg-fte.tex}}
    \label{tab:hiring-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}: 
    This table show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on the number of faculty hires (per student) total for 2011--2021 at US public universities, using the funding shock to instrument for state funding.
    Each observation is a university--year at a public university, where funding data come from IPEDS and faculty count total for 2011--2021 from \citep{wapman2022quantifying}.
    The panels show the effect of a $1$\% change in state funding per student at the university (total for 2011--2021) on the number of new faculty hires by gender (and all).
    Standard errors are clustered at the university-year level, and university $+$ year fixed effects are included through--out.
\end{table}

@@

The implied marginal rate of substitution between lecturers and assistant professors is estimated as -3.26 (standard error 0.50), based on 10,000 bootstrap samples.
This means that public universities increased their number lecturers per student by 3.26\% when they decreased their count of assistant professors, on average and subject to the changes in state funding they experienced 1990--2017.
Between lecturers and full professors the rate of substitution is -3.19 (0.34), which implies that universities substitute between lecturers and full professors in the same way.
Between assistant and full professors the rate of substitution is 0.99 (0.11), which intuitively implies that universities treated assistant and full professors (i.e., those before and after tenure in the tenure system) as complements.
